---
title: "SOM 3 LARVAS"
output: html_notebook
---

#1. Loading raw data and getting it ready

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(readr)
larva3 <- read_csv("~/GENER/SOMs/Ricardo/1. Raw Data/PSOMML_3Larvas.csv", 
    col_names = FALSE)
colnames(larva3) <- c("FILA", "COLUMNA", "RED", "GREEN", "BLUE")

larva3 <- larva3[,-1:-2]


```

The dataframe obtained is made of `integer` values:

```{r}
str(larva3)
```


This will be a problem down the line when feeding this data to the algorithm, values have to be changed to  `numeric`:

```{r}
larva3[,'RED'] <- as.numeric(as.factor(larva3$RED))
larva3[,'GREEN'] <- as.numeric(as.factor(larva3$GREEN))
larva3[,'BLUE'] <- as.numeric(as.factor(larva3$BLUE))
```

Whit this taken care of we have a database ready to be fed to the algorithmL
```{r}
str(larva3)
```
#2 Initial exploration of data

A brief summary of the values of the variable can help with the initial understanding of things:
```{r}
summary(larva3)
```


#3 Self Organizing Map

##Loading required package

```{r, message=FALSE, warning=FALSE}
require(kohonen)
```

##Changing dataframe to a `matrix`
```{r}
#I removed the function scale() because it does not need to be scaled.
larva3.matrix <- as.matrix(larva3)
```

##Creating the SOM grid
```{r}
som_grid3 <- somgrid(xdim = 40, ydim = 40, topo = "hexagonal")
```


##Training the SOM
```{r, message=FALSE, warning=FALSE, include=FALSE}
som_model3 <- som(larva3.matrix,
                 grid = som_grid,
                 rlen = 1500,
                 alpha = c(0.05,0.01),
                 keep.data = TRUE)
```

##Training progress evaluation
```{r, message=FALSE, warning=FALSE}
plot(som_model3, type = "changes")
```

##Node counts
```{r, fig.height= 8}
source('coolBlueHotRed.R')
plot(som_model3, type = "counts", palette.name=coolBlueHotRed)
```

##SOM by RGB value

```{r, fig.height=10}
par(mfrow = c(2,2))
models3 <- as.data.frame(som_model3$codes)
plot(som_model3,
     type = "property",
     property = models3[,1],
     main=names(models3)[1],
     palette.name=coolBlueHotRed)
plot(som_model3,
     type = "property",
     property = models3[,2],
     main=names(models3)[2],
     palette.name=coolBlueHotRed)
plot(som_model3,
     type = "property",
     property = models3[,3],
     main=names(models3)[3],
     palette.name=coolBlueHotRed)

```


##Neighbor Distance
```{r, fig.align='center'}
plot(som_model3, type = "dist.neighbours", palette.name = grey.colors)
```

##Code Spread
```{r, fig.height= 10}
plot(som_model3, type = "codes", palette.name = coolBlueHotRed)
```

##Clustering
```{r}
mydata <- som_model3$codes 
wss <- (nrow(mydata)-1)*sum(apply(data.frame(mydata),2,var)) 
for (i in 2:20) {
  wss[i] <- sum(kmeans(data.frame(mydata), centers=i)$withinss)
}
plot(1:20, wss, type="b", xlab="Number of Clusters",
     ylab="Within groups sum of squares", main="Within cluster sum of squares (WCSS)")

som_cluster <- cutree(hclust(dist(data.frame(som_model3$codes))), 3)


```

In this graphic we can see that the ideal amount of clusters is not more than 5. Starting with 5 and going down we will choose the smallest number that gives us a smooth distribution in quantity of it's members:

```{r}
# 7 Clusters -- the first cluster has considerably less members than the other four ones.
table(cutree(hclust(dist(data.frame(som_model3$codes))), 3))


# 5 Clusters -- the first cluster has considerably less members than the other four ones.
table(cutree(hclust(dist(data.frame(som_model3$codes))), 5))

#4 clusters -- we have founde the good one.
table(cutree(hclust(dist(data.frame(som_model3$codes))), 4))
```


```{r, fig.height= 10}
pretty_palette <- c("#1f77b4", '#ff7f0e', '#2ca02c', '#d62728', '#9467bd', '#8c564b', '#e377c2')

plot(som_model3, type="mapping", bgcol = pretty_palette[som_cluster], main = "Clusters", keepMargins = TRUE) 
add.cluster.boundaries(som_model3, som_cluster, lwd = 7)
```

