---
title: "SOM for Ricardo's Data"
output: html_notebook
---

#1. Loading raw data and getting it ready

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(readr)
img <- read_csv("~/Documentos/GENER/ricardo/1. Raw Data/PSOMMLmini.csv", 
    col_names = FALSE)
colnames(img) <- c("FILA", "COLUMNA", "RED", "GREEN", "BLUE")
```

The dataframe obtained is made of `integer` values:

```{r}
str(img)
```


This will be a problem down the line when feeding this data to the algorithm, values have to be changed to  `numeric`:

```{r}
img[,'FILA'] <- as.numeric(as.factor(img$FILA))
img[,'COLUMNA'] <- as.numeric(as.factor(img$COLUMNA))
img[,'RED'] <- as.numeric(as.factor(img$RED))
img[,'GREEN'] <- as.numeric(as.factor(img$GREEN))
img[,'BLUE'] <- as.numeric(as.factor(img$BLUE))
```

Whit this taken care of we have a database ready to be fed to the algorithmL
```{r}
str(img)
```
#2 Initial exploration of data

A brief summary of the values of the variable can help with the initial understanding of things:
```{r}
summary(img)
```


#3 Self Organizing Map

##Loading required package

```{r, message=FALSE, warning=FALSE}
require(kohonen)
```

##Changing dataframe to a `matrix` and normalizing it's values
```{r}
#I removed the function scale() because it does not need to be scaled.
img.matrix <- as.matrix(img)

#TEST
img.matrixTest <- img.matrix[1:20000,]
```

##Creating the SOM grid
```{r}
som_grid <- somgrid(xdim = 70, ydim = 70, topo = "rectangular")
```


##Training the SOM
```{r, message=FALSE, warning=FALSE, include=FALSE}
som_model <- som(img.matrixTest,
                 grid = som_grid,
                 rlen = 200,
                 alpha = c(0.05,0.01),
                 keep.data = TRUE)
```

##Training progress evaluation
```{r}
plot(som_model, type = "changes")
```

##Node counts
```{r}
source('coolBlueHotRed.R')
plot(som_model, type = "counts", palette.name=coolBlueHotRed)
```

##Neighbor Distance
```{r}
plot(som_model, type = "dist.neighbours", palette.name = grey.colors)
```

##Code Spread
```{r}
plot(som_model, type = "codes", palette.name = coolBlueHotRed)
```

##Clustering
```{r}
mydata <- som_model$codes 
wss <- (nrow(mydata)-1)*sum(apply(data.frame(mydata),2,var)) 
for (i in 2:15) {
  wss[i] <- sum(kmeans(data.frame(mydata), centers=i)$withinss)
}
plot(wss)

som_cluster <- cutree(hclust(dist(data.frame(som_model$codes))), 5)
```

```{r}
pretty_palette <- c("#1f77b4", '#ff7f0e', '#2ca02c', '#d62728', '#9467bd', '#8c564b', '#e377c2')

plot(som_model, type="mapping", bgcol = pretty_palette[som_cluster], main = "Clusters") 
add.cluster.boundaries(som_model, som_cluster)
```

